Natural selection in contemporary humans is linked to income and substitution effects

### 2020-12-21 why-natural-selection.Rmd
Last tweaks. (I hope.)

---
title: "Labor Market Income and Substitution Effects Explain Contemporary Natural Selection"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores which predict lower (higher) earnings and education are selected for (against). Selection effects are concentrated among lower-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. We find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects and that are strongly related to the socio-economic system."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

We created polygenic scores for 33 traits in `r nrow(famhist)` individuals,
corrected for ancestry using 100 genetic principal components (see Materials and
Methods). Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores in
the sample by 5-year birth intervals. Several scores show consistent increases
or declines over this 30-year period, of the order of 5% of a standard
deviation. These changes could reflect natural selection within the UK
population, but also emigration, or ascertainment bias within the sample.
Respondents have higher income and are better educated than the UK population,
and they may also differ on other unobserved characteristics
[@10.1093/aje/kwx246]. Since richer and educated people also live longer, this
bias could also increase with age.

@@

```{r plot-income-educ-level, fig.cap = "Selection effects by education and income. Each point represents a single bivariate regression of number of children on a polygenic score", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}

@@

```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth", fig.subcap = c("Lifetime number of sexual partners", "Household type", "Age at first live birth, females"), fig.ncol = 1, fig.align = "center"}

@@

To investigate this further, we regress *number of children* on polygenic scores
*controlling* for AFLB (Figure \@ref(fig:plot-age-flb)b). In `r opp_signed` out
of `r length(score_names)` cases, effects change sign when controls are added.
The correlation between effect sizes controlling for AFLB, and raw effect sizes,
is `r cor_raw_flb`. Thus, selection effects seem to come through two opposing
channels: an effect on AFLB, and an opposite-signed effect on number of children
controlling for AFLB. Note that we do not claim that AFLB is exogenous to an
individual's polygenic scores -- indeed, our analyses show that it is not
(Appendix Figure \@ref(fig:plot-age-flb-dv)). Rather, we argue that AFLB mediates
part of the relationship between polygenic scores and fertility, and that
once this is controlled for, the remaining part of the relationship has the
opposite sign.

@@

An alternative theory is that traits selected for are linked to externalizing
behaviour, risk-seeking and low time discount rates, via the channel of early
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and (younger) age at smoking initiation, are selected for. However,
this theory is less good at explaining variation in selection across the full
range of scores, including physical measures, e.g. waist-hip ratio and BMI.
Externalizing behaviour also does not explain why selection effects should work
in the opposite direction among older parents, whereas the economic theory can
explain this via income effects, as described above. We test the alternative
theory directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040). While risk attitude is always
a highly significant predictor of number of children, it has little impact on
the effect size or significance of the polygenic scores. The median ratio of
effect sizes between regressions with and without controls is `r median(ratio)`;
all scores which are significant at $p \le 0.05/33$ in uncontrolled regressions
remain so when controlling for risk attitude. Thus, although risk attitude does
predict number of children, it appears not to mediate selection effects.
Overall, we believe that the economic theory is the most likely explanation.

@@

Previous work has documented natural selection in modern populations on variants
underlying polygenic traits [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that correlations between polygenic scores and
fertility are highly concentrated among specific subgroups of the population,
including people with lower income, lower education, younger first parenthood, 
and more lifetime sexual partners. Indeed, among older parents, selection effects are
actually reversed. Furthermore, the size of selection effects on a polygenic
score correlates with that score's association with labour market earnings. The
economic theory of fertility provides a parsimonious explanation for this.

@@

Any model of fertility is implicitly a model of natural selection, but so far,
the economic and human genetics literatures have developed in parallel.
Integrating the two could deepen our understanding of natural selection in
modern societies. Economics possesses a range of theoretical models on the
effects of skills, education and income [see @hotz1997economics;
@lundberg2007american]. One perennial problem is how to test these theories in a
world where education, labour and marriage markets all interact. Genetic data,
such as polygenic scores, could help to pin down the direction of causality.
Conversely, theory and empirical results from economics can shine a light on the
mechanisms behind natural selection, and thereby on the nature of individual
differences in complex traits and disease risk.

@@

```{r plot-sexes, fig.cap = "Selection effects by sex. Lines show significant differences."}

@@

```{r plot-purifying, fig.align = "center", fig.cap = "Purifying/diversifying selection: predicted number of children by polygenic score", fig.height = 5}

@@

```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by Townsend deprivation quintile of birth area", fig.align = "center"}

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term     = pretty_names(term),
        term     = reorder_within(term, estimate, dep.var, order_abs(1)),
        dep.var  = fct_inorder(dep.var),
        reg.type = fct_recode(reg.type, 
                               Residualized   = "controlled", 
                               Unresidualized = "raw"
                             )
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = term, x = estimate, fill = reg.type, 
        colour = p < 0.05/{{n_regs}})) +
        my_vline + 
        facet_wrap(vars(dep.var), scales = "free") +
        pval_geom_point(position = position_dodge(0.4), alpha = 0.8) +
        my_labs() +
        pval_scale() +
        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
        my_theme() +
        scale_y_reordered() +
        theme(
          axis.text.x = element_text(size = 8)
        )

@@

Another way to examine the "earnings" theory of natural selection is to compare
selection effects of PGS with their genetic correlation with educational
attainment (EA3). Since EA3 strongly predicts earnings, if earnings drives
differences in fertility, we'd expect a correlation between the two sets of
results. Figure \@ref(fig:plot-rgs-by-effect-size) shows this is so. Genetic
correlations were calculated using LD score regression from GWAS summary
statistics.

@@

rgs %>% 
      filter(p2 != "EA3_excl_23andMe_UK", p2 != "EA2_noUKB") %>% 
      ggplot(aes(rg, estimate)) + 
        geom_point() +
        my_vline +
        my_hline +
        labs(x = "Genetic correlation with EA3", y = "Effect on number of children")

@@

@@

ineq_comp %>% 
      select(
        PGS                 = score, 
        `Without selection` = diff_hypoth,
        `With selection`    = diff_children, 
        `% change`          = pct_change
      ) %>% 
      mutate(
        PGS = pretty_names(PGS)
      ) %>% 
      arrange(desc(`% change`)) %>% 
      as_hux() %>% 
      set_align(everywhere, -1, "right") %>% 
      set_number_format(everywhere, 2:3, 3) %>% 
      set_number_format(everywhere, 4, 1) %>% 
      set_caption("Differences in polygenic scores between highest and lowest income group.") %>% 
      theme_compact() %>% 
      set_width(0.9) %>% 
      set_col_width(c(.35, .15, .15 , .15)) %>% 
      set_bottom_border(final(1), everywhere)
```

>>>>

---
title: "Natural selection in contemporary humans is linked to income and substitution effects"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores which predict lower (higher) earnings and education are selected for (against). Selection effects are concentrated among lower-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner. Effect sizes are substantially larger after correcting for ascertainment bias. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market. We find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects and that are strongly related to the socio-economic system."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
individuals, corrected for ancestry using 100 genetic principal components (see
Materials and Methods). Figure \@ref(fig:plot-means-over-time) plots mean
polygenic scores in the sample by 5-year birth intervals. Several scores show
consistent increases or declines over this 30-year period, of the order of 5% of
a standard deviation. These changes could reflect natural selection within the
UK population, but also emigration, or ascertainment bias within the sample.
Respondents have higher income and are better educated than the UK population,
and they may also differ on other unobserved characteristics
[@10.1093/aje/kwx246]. Since richer and educated people also live longer, this
bias could also increase with age.

@@

```{r plot-income-educ-level, fig.cap = "Selection effects by education and income. Each point represents a single bivariate regression of number of children on a polygenic score.", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}

@@

```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth.", fig.subcap = c("Lifetime number of sexual partners", "Household type", "Age at first live birth, females"), fig.ncol = 1, fig.align = "center"}

@@

To investigate this further, we regress *number of children* on polygenic scores
*controlling* for AFLB, again among females (Figure \@ref(fig:plot-age-flb)b).
In `r opp_signed` out of `r length(score_names)` cases, effects change sign when
controls are added. The correlation between effect sizes controlling for AFLB,
and raw effect sizes, is `r cor_raw_flb`. Thus, selection effects seem to come
through two opposing channels: an effect on AFLB, and an opposite-signed effect
on number of children controlling for AFLB. Note that we do not claim that AFLB
is exogenous to an individual's polygenic scores -- indeed, our analyses show
that it is not (Appendix Figure \@ref(fig:plot-age-flb-dv)). Rather, we argue
that AFLB mediates part of the relationship between polygenic scores and
fertility, and that once this is controlled for, the remaining part of the
relationship has the opposite sign.

@@

An alternative theory is that traits selected for are linked to externalizing
behaviour, risk-seeking and low time discount rates, via the channel of early
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and (younger) age at smoking initiation, are selected for. However,
this theory is less good at explaining variation in selection across the full
range of scores, including physical measures, e.g. waist-hip ratio and BMI.
Externalizing behaviour also does not explain why selection effects should work
in the opposite direction among older mothers, whereas the economic theory can
explain this via income effects, as described above. We test the alternative
theory directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040). While risk attitude is always
a highly significant predictor of number of children, it has little impact on
the effect size or significance of the polygenic scores. The median ratio of
effect sizes between regressions with and without controls is `r median(ratio)`;
all scores which are significant at $p \le 0.05/33$ in uncontrolled regressions
remain so when controlling for risk attitude. Thus, although risk attitude does
predict number of children, it appears not to mediate selection effects.
Overall, we believe that the economic theory is the most likely explanation.

@@

Previous work has documented natural selection in modern populations on variants
underlying polygenic traits [@beauchamp2016genetic; @kong2017selection;
@sanjak2018evidence]. We show that correlations between polygenic scores and
fertility are highly concentrated among specific subgroups of the population,
including people with lower income, lower education, younger first parenthood, 
and more lifetime sexual partners. Indeed, among older mothers, selection effects are
actually reversed. Furthermore, the size of selection effects on a polygenic
score correlates with that score's association with labour market earnings. The
economic theory of fertility provides a parsimonious explanation for this.

@@

Any model of fertility is implicitly a model of natural selection, but so far,
the economic and human genetics literatures have developed in parallel.
Integrating the two could deepen our understanding of natural selection in
modern societies. Economics possesses a range of theoretical models on the
effects of skills, education and income [see @hotz1997economics;
@lundberg2007american]. One perennial problem is how to test these theories in a
world where education, labour and marriage markets all interact. Genetic data,
such as polygenic scores, could help to pin down the direction of causality,
for example via Mendelian randomization [@davey2003mendelian].
Conversely, theory and empirical results from economics can shine a light on the
mechanisms behind natural selection, and thereby on the nature of individual
differences in complex traits and disease risk.

@@

```{r plot-sexes, fig.cap = "Selection effects by sex. Highlighted lines are significant differences. Highlighted points are significantly different from 0."}

@@

```{r plot-purifying, fig.align = "center", fig.cap = "Purifying/diversifying selection: predicted number of children by polygenic score.", fig.height = 5}

@@

```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by Townsend deprivation quintile of birth area.", fig.align = "center"}

@@

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term     = pretty_names(term),
        term     = reorder_within(term, estimate, dep.var, order_abs(1)),
        dep.var  = fct_inorder(dep.var),
        reg.type = fct_recode(reg.type, 
                               `Resid. for 100 PCs`  = "controlled", 
                               Unresidualized        = "raw"
                             )
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(y = term, x = estimate, fill = reg.type, 
        colour = p < 0.05/{{n_regs}})) +
        my_vline + 
        facet_wrap(vars(dep.var), scales = "free") +
        pval_geom_point(position = position_dodge(0.4), alpha = 0.8) +
        my_labs() +
        pval_scale() +
        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
        my_theme() +
        scale_y_reordered() +
        theme(
          axis.text.x  = element_text(size = 7),
          legend.text  = element_text(size = 8),
          legend.title = element_text(size = 9)
        )

@@

@@

rgs_cor <- cor.test(~ rg + estimate, 
                      data = rgs %>%
                               filter(
                                 p2 != "EA3_excl_23andMe_UK", 
                                 p2 != "EA2_noUKB"
                               )
                    )
rgs %>% 
      filter(p2 != "EA3_excl_23andMe_UK", p2 != "EA2_noUKB") %>% 
      ggplot(aes(rg, estimate)) + 
        geom_point() +
        my_vline +
        my_hline +
        labs(x = "Genetic correlation with EA3", y = "Effect on number of children")

@@

Another way to examine the "earnings" theory of natural selection is to compare
selection effects of polygenic scores with their genetic correlation with educational
attainment (EA3). Since EA3 strongly predicts earnings, if earnings drives
differences in fertility, we'd expect a correlation between the two sets of
results. Figure \@ref(fig:plot-rgs-by-effect-size) shows this is so: the 
correlation is `r rgs_cor$estimate`. Genetic
correlations were calculated using LD score regression from GWAS summary
statistics.

@@

ineq_comp %>% 
      select(
        PGS                 = score, 
        `Without selection` = diff_hypoth,
        `With selection`    = diff_children, 
        `% change`          = pct_change
      ) %>% 
      mutate(
        PGS = pretty_names(PGS)
      ) %>% 
      arrange(desc(`% change`)) %>% 
      as_hux() %>% 
      set_align(everywhere, -1, "right") %>% 
      set_number_format(everywhere, 2:3, 3) %>% 
      set_number_format(everywhere, 4, 1) %>% 
      set_caption("Differences in polygenic scores between highest and lowest income group.") %>% 
      theme_compact() %>% 
      set_width(0.75) %>% 
      set_col_width(c(.4, .2, .2, .2)) %>% 
      set_bottom_border(final(1), everywhere) %>% 
      set_position("center")
```
